Changepoint Approach

### 2022-07-25 appendices/changepoint_methods.Rmd
complete changepoint approach appendix

The aim of this appendix is to present results for an alternative method of attempting to detect discontinuities in the series of life expectancy changes for different countries.

@@

```{r}
# load packages
pacman::p_load(here, tidyverse)

# load data 
hmd_lt <- read_rds(here("data", "lifetables.rds"))

@@

source(here("R", "make_synthetic_germany_functions.R"))
source(here("R", "make_pop_selection.R"))

@@

```{r}
change_in_ex_selected_countries %>% 
  filter(x == 0) %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  filter(!is.na(country)) %>% 
  filter(between(year, 1980, 2020)) %>% 
  ggplot(aes(x = year, y = delta_ex)) + 
  geom_point() + 
  facet_grid(sex~country) + 
  geom_hline(yintercept = 0) + 
  scale_y_continuous(breaks = seq(-30, 50, by = 10)) +
  theme(
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
  ) + 
  labs(
    x = "Year",
    y = "Change in life expectancy from previous year",
    title = "Annual change in life expectancy at birth, selected countries",
    subtitle = "Line: nonlinear smoother over the points",
    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
  )

change_in_ex_selected_countries %>% 
  filter(x == 65) %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  filter(!is.na(country)) %>% 
  filter(between(year, 1980, 2020)) %>% 
  ggplot(aes(x = year, y = delta_ex)) + 
  geom_point() + 
  facet_grid(sex~country) + 
  geom_hline(yintercept = 0) + 
  scale_y_continuous(breaks = seq(-30, 50, by = 10)) +
  theme(
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
  ) + 
  labs(
    x = "Year",
    y = "Change in life expectancy from previous year",
    title = "Annual change in life expectancy at age 65, selected countries",
    subtitle = "Line: nonlinear smoother over the points",
    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
  )

@@

@@

@@

(The optim-based function takes far too long to run, so is not used)

@@

```{r}
changepoint_breakpoint_models <- 
  hmd_ex_selected_countries_with_synth %>% 
  filter(code != "DEUTNP") %>%   
  filter(year >= 1979) %>% 
  group_by(code, x, sex) %>% 
  nest() %>% 
  mutate(
    mdl_outputs = map(data, get_best_tau_from_gridsearch, what = "all")
  )
```
The relationship between proposed `tau` (i.e. changepoint year) and BIC is shown for England & Wales below

```{r}
changepoint_breakpoint_models %>% 
  filter(code == "GBRTENW") %>% 
  unnest(mdl_outputs) %>% 
  group_by(x, sex) %>% 
  mutate(min_bic = bic == min(bic)) %>% 
  ungroup() %>% 
  ggplot(aes(tau, bic, colour = min_bic, shape = min_bic)) + 
  geom_point() + 
  facet_grid(x ~ sex) +
  scale_colour_manual(values = c(`TRUE` = 'red', `FALSE` = "black")) + 
  scale_shape_manual(values = c(`TRUE` = 'triangle', `FALSE` = 'circle'))

@@

>>>>

The aim of this appendix is to present results for an alternative method of attempting to detect discontinuities in the series of life expectancy changes for different countries. The main approach used was segmented regression using the `segmented` package. The segmented regressions were fit to the data itself $y ~ t$. This alternative approach instead involves fitting data to the first differences $y(t) - y(t-1) ~ t$.

The alternative approach identified some similar breakpoints/changepoints to the main, segmented approach. However it differed for many populations in identifying adjacent or near-adjacent changepoints in the data. Further inspection of these pairs of changepoints indicated that the approach developed inadvertently operated as a form of outlier detector algorithm, removing from the series extreme values. Though extreme values are a genuine phenomenon within mortality change data, they were not the primary purpose of the analysis, and so this approach was not used within the main manuscript.

@@

```{r}
# load packages

library(tidyverse)

# load data 
hmd_lt <- read_rds("https://github.com/JonMinton/change-in-ex/blob/main/data/lifetables.rds?raw=true")

@@

source("https://raw.githubusercontent.com/JonMinton/change-in-ex/main/R/make_synthetic_germany_functions.R")

source("https://raw.githubusercontent.com/JonMinton/change-in-ex/main/R/make_pop_selection.R")

@@

```{r}
change_in_ex_selected_countries %>% 
  filter(x == 0) %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  filter(!is.na(country)) %>% 
  filter(between(year, 1980, 2020)) %>% 
  ggplot(aes(x = year, y = delta_ex)) + 
  geom_point() + 
  facet_grid(sex~country) + 
  geom_hline(yintercept = 0) + 
  scale_y_continuous(breaks = seq(-30, 50, by = 10)) +
  theme(
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
  ) + 
  labs(
    x = "Year",
    y = "Change in life expectancy from previous year",
    title = "Annual change in life expectancy at birth, selected countries",
    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
  )

change_in_ex_selected_countries %>% 
  filter(x == 65) %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  filter(!is.na(country)) %>% 
  filter(between(year, 1980, 2020)) %>% 
  ggplot(aes(x = year, y = delta_ex)) + 
  geom_point() + 
  facet_grid(sex~country) + 
  geom_hline(yintercept = 0) + 
  scale_y_continuous(breaks = seq(-30, 50, by = 10)) +
  theme(
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
  ) + 
  labs(
    x = "Year",
    y = "Change in life expectancy from previous year",
    title = "Annual change in life expectancy at age 65, selected countries",
    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
  )

```
# Functions

@@

# This function was not used in the end. The reason is because the fitness landscape is 
# difficult to navigate. Instead a gridsearch approach was used

@@

# This is the function actually used, for a single changepoint value

@@

# Running the functions

## Single changepoint

@@

```{r}
changepoint_breakpoint_models <- 
  hmd_ex_selected_countries_with_synth %>% 
  filter(code != "DEUTNP") %>%   
  filter(year >= 1979) %>% 
  group_by(code, x, sex) %>% 
  nest() %>% 
  mutate(
    mdl_outputs = map(data, get_best_tau_from_gridsearch, what = "all")
  )
```

The relationship between proposed `tau` (i.e. changepoint year) and BIC is shown for England & Wales below

```{r}
changepoint_breakpoint_models %>% 
  filter(code == "GBRTENW") %>% 
  unnest(mdl_outputs) %>% 
  group_by(x, sex) %>% 
  mutate(min_bic = bic == min(bic)) %>% 
  ungroup() %>% 
  ggplot(aes(tau, bic, colour = min_bic, shape = min_bic)) + 
  geom_point() + 
  facet_wrap(~ sex + x, scales = "free_y") +
  scale_colour_manual(values = c(`TRUE` = 'red', `FALSE` = "black")) + 
  scale_shape_manual(values = c(`TRUE` = 'triangle', `FALSE` = 'circle')) + 
  labs(
    title = "BIC by proposed changepoint year",
    subtitle = "Red triangle = lowest BIC. Population: England & Wales",
    x = expression(paste("Proposed changepoint year ", tau)),
    y = "BIC (lower is better)",
    caption = "Source: HMD Lifetables"
  ) + 
  theme(legend.position = "none")

@@

We can see here that 2012 is consistently identified as having the lowest BIC for each of the four subpopulations, except for males where x = 0, where 2013 is identified instead. We can also see that the 'fitness landscape' (i.e. how BIC changes as the proposed changepoing years change) is quite uneven, and so simple numerical optimisation algorithms like Newton-Ralphson are likely to get 'stuck' at local rather than the global optima, and so the results of such algorithms may be quite dependent on starting parameters.

## For two changepoints

The following code identifies the best two changepoint model for each population

```{r}
changepoint_breakpoints_models <- 
  hmd_ex_selected_countries_with_synth %>% 
  filter(code != "DEUTNP") %>%   
  filter(year >= 1979) %>% 
  group_by(code, x, sex) %>% 
  nest() %>% 
  mutate(
    mdl_outputs = map(data, get_best_taus_from_gridsearch, what = "all")
  )

Whereas for a single changepoint the relationship between BIC and proposed $\tau$ can be visualised as a scatterplot, for two values $\tau_1$ and $\tau_2$ they can be visualised as a heatmap instead. Our only known constraint is that $\tau_2 > \tau_1$, meaning within the heatmap only a triangle rather than rectangle of candidate values needs to be considered.

The following shows this heatmap for England & Wales. A red dot is added to indicate the pair of candidate values $\{\tau_1, \tau_2\}$ with the lowest BIC. As the range of BIC differs with each dataset, the values are normalised for each subpopulations such that the highest value is 1 and the lowest 0.

```{r}
changepoint_breakpoints_models %>% 
  filter(code == "GBRTENW") %>% 
  unnest(mdl_outputs) %>% 
  group_by(x, sex) %>% 
  mutate(bic_std = (bic - min(bic)) / (max(bic) - min(bic))) %>% 
  ggplot(aes(x = tau1, y = tau2, colour = bic_std, fill = bic_std)) + 
  geom_raster() + 
  facet_wrap(~ sex + x, ncol = 2) + 
  scale_color_viridis_c() + 
  geom_point(colour= 'red', 
             data = changepoint_breakpoints_models %>% 
              filter(code == "GBRTENW") %>% 
              unnest(mdl_outputs) %>% 
              group_by(x, sex) %>% 
              mutate(bic_std = (bic - min(bic)) / (max(bic) - min(bic))) %>% 
              filter(bic_std == min(bic_std))
  ) + 
  coord_fixed() +
  theme(axis.text.x = element_text(angle = 90)) + 
  labs(
    x = expression(paste("Proposed ", tau[1])),
    y = expression(paste("Proposed ", tau[2])),
    fill = "Standardised BIC",
    title = "The relationship between BIC and\nproposed changepoint values",
    subtitle = "Population: England & Wales. Red dot: pair of values with lowest BIC",
    caption = "Source: HMD lifetables"
  )
```

We can see in the above that, for males for x = 65, a $\{\tau_1, \tau_2\}$ pairing of `{2000, 2012}` has been identified. For males and females where x = 0, a near-contiguous pairing of `{2009, 2012}` has been identified. And for females where x = 65 a contigious pairing of `{2015, 2016}` had the lowest BIC.

Similar identification of contiguous and near-contiguous pairings were also identified for many other countries. For example, the following shows this for France.

```{r}
changepoint_breakpoints_models %>% 
  filter(code == "FRATNP") %>% 
  unnest(mdl_outputs) %>% 
  group_by(x, sex) %>% 
  mutate(bic_std = (bic - min(bic)) / (max(bic) - min(bic))) %>% 
  ggplot(aes(x = tau1, y = tau2, colour = bic_std, fill = bic_std)) + 
  geom_raster() + 
  facet_wrap(~ sex + x, ncol = 2) + 
  scale_color_viridis_c() + 
  geom_point(colour= 'red', 
             data = changepoint_breakpoints_models %>% 
              filter(code == "FRATNP") %>% 
              unnest(mdl_outputs) %>% 
              group_by(x, sex) %>% 
              mutate(bic_std = (bic - min(bic)) / (max(bic) - min(bic))) %>% 
              filter(bic_std == min(bic_std))
  ) + 
  coord_fixed() +
  theme(axis.text.x = element_text(angle = 90)) + 
  labs(
    x = expression(paste("Proposed ", tau[1])),
    y = expression(paste("Proposed ", tau[2])),
    fill = "Standardised BIC",
    title = "The relationship between BIC and\nproposed changepoint values",
    subtitle = "Population: FRATNP. Red dot: pair of values with lowest BIC",
    caption = "Source: HMD lifetables"
  )

```
For France, for all four subpopulations, a pairing of `{2004, 2005}` was identified. If we look at the series for France, we can see why this pair of years may have been identified

```{r}
change_in_ex_selected_countries %>% 
  filter(code == "FRATNP") %>%   
  filter(year >= 1979) %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("France"))) %>% 
  filter(!is.na(country)) %>% 
  filter(between(year, 1980, 2020)) %>% 
  mutate(is_2004 = year == 2004) %>% 
  ggplot(aes(x = year, y = delta_ex)) + 
  geom_point(aes(shape = is_2004, colour = is_2004)) + 
  facet_grid(sex ~ x) +
  scale_shape_manual(values =c(`TRUE` = "triangle", `FALSE` = "circle")) + 
  scale_colour_manual(values =c(`TRUE` = "darkred", `FALSE` = "black")) + 
  geom_hline(yintercept = 0) + 
  theme(
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        legend.position = 'none'
  ) + 
  labs(
    x = "Year",
    y = "Change in life expectancy from previous year",
    title = "Annual change in life expectancy in France",
    subtitle = "Columns: starting age; Red triangle: Year 2004",
    caption =  "Source: Human Mortality Database"
  )

We can see that for each of this subpopulation, year = 2004 (i.e. the change in ex from 2003 to 2004) was much higher than all other years, and so these points are outliers. So, it appears the changepoint approach has inadertently 'detected' this outlier.

# Best model for each population

The following code shows the best performing model, from the best zero, one and two changepoint models, for each of the populations.

```{r}
estimate_null <- function(df){
  df %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
    filter(!is.na(delta_ex)) %>% 
    lm(delta_ex ~ 1, data = .)
}

  hmd_ex_selected_countries_with_synth %>% 
    filter(code != "DEUTNP") %>%   
    filter(year >= 1979) %>% 
    group_by(code, x, sex) %>% 
    nest() %>% 
    mutate(
      mdl_0 = map(data, estimate_null),
      bic_0    = map_dbl(mdl_0, BIC)
    ) %>% 
    select(code, x, sex, mdl_0, bic_0) %>% 
    left_join(
      changepoint_breakpoint_models %>% 
        unnest(mdl_outputs) %>% 
        group_by(code, x, sex) %>% 
        filter(bic == min(bic)) %>% 
        select(code, x, sex, tau1 = tau, bic_1 = bic)  
    ) %>% 
    left_join(
      changepoint_breakpoints_models %>% 
        unnest(mdl_outputs) %>% 
        group_by(code, x, sex) %>% 
        filter(bic == min(bic)) %>% 
        select(code, x, sex, tau1_2 = tau1, tau2_2 = tau2, bic_2 = bic)
    ) %>% 
  select(-mdl_0) %>% 
  ungroup() # %>% 
run_tau_1 <- function(tau, df){
  df %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
    filter(!is.na(delta_ex)) %>% 
    mutate(T_param = ifelse(year < tau, FALSE, TRUE)) %>% 
    lm(delta_ex ~ T_param, data = .)
}

run_tau_2 <- function(tau1, tau2, df){
  df %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
    filter(!is.na(delta_ex)) %>% 
    mutate(T_param = ifelse(
      year < tau1, 
      "bp0", 
      ifelse(
        year < tau2, "bp1", "bp2"
        )
      )
    ) %>% 
    lm(delta_ex ~ T_param, data = .)
}

make_path_coords <- function(df, taus = NULL, mdl){
  stopifnot(
    "wrong number of coefficients for the model type" = 
    length(coefficients(mdl)) == length(taus) + 1
  )

  min_x = min(df$year)
  max_x = max(df$year)

  coeffs <- coefficients(mdl)

  y0 <- coeffs[1]

  out <- tribble(
    ~x, ~y,
    min_x, y0
  )

  if(length(coeffs) == 1){
    out <- out %>%
      bind_rows(
        tribble(
          ~x, ~y,
          max_x, y0
        )
      )
    return(out)

  } else if (length(coeffs) == 2){
    out <- out %>% 
      bind_rows(
        tribble(
          ~x, ~y,
          taus[1], y0,
          taus[1], y0 + coeffs[2],
          max_x, y0 + coeffs[2]
        )
      )
    return(out)

  } else if (length(coeffs) == 3){
    out <- out %>% 
      bind_rows(
        tribble(
          ~x, ~y,
          taus[1], y0,
          taus[1], y0 + coeffs[2],
          taus[2], y0 + coeffs[2],
          taus[2], y0 + coeffs[3],
          max_x, y0 + coeffs[3]

        )
      )

    return(out)
  }
  return(NULL) # should not be triggered  
}

# Let's try this with 0cp

paths_0cp <- 
  hmd_ex_selected_countries_with_synth %>% 
    filter(code != "DEUTNP") %>%   
    filter(year >= 1979) %>% 
    group_by(code, x, sex) %>% 
    nest() %>% 
    mutate(
      mdl_0 = map(data, estimate_null)
    ) %>% 
    mutate(bic = map_dbl(mdl_0, BIC)) %>% 
    mutate(
      tau_paths = pmap(.l = list(df = data, mdl = mdl_0), make_path_coords)
    ) %>% 
  select(code, x, sex, bic_0cp = bic, paths_0cp = tau_paths)

# Now 1cp
paths_1cp <- 
  changepoint_breakpoint_models %>% 
    unnest(mdl_outputs) %>% 
    group_by(code, x, sex) %>% 
    filter(bic == min(bic)) %>% 
    ungroup() %>% 
    mutate(
      mdl_with_best_1cp = map2(tau, data, run_tau_1)
    ) %>% 
    mutate(
      tau_paths = pmap(
        .l = list(
          df = data, 
          mdl = mdl_with_best_1cp, 
          taus = c(tau)
        ), 
        make_path_coords
      )
    ) %>% 
    select(
      code, x, sex, bic_1cp = bic, paths_1cp = tau_paths
    )

# and now for 2cp
paths_2cp <- 
  changepoint_breakpoints_models %>% 
    unnest(mdl_outputs) %>% 
    group_by(code, x, sex) %>% 
    filter(bic == min(bic)) %>% 
    ungroup() %>% 
    mutate(mdl_with_best_2cp = pmap(.l = list(tau1=tau1, tau2=tau2, df=data), run_tau_2)) %>%
    mutate(taus = map2(tau1, tau2, c)) %>% 
    mutate(
      tau_paths = pmap(
        .l = list(
          df = data, 
          mdl = mdl_with_best_2cp, 
          taus = taus
        ), 
        make_path_coords
      )
    ) %>% 
    select(code, x, sex, bic_2cp = bic, paths_2cp = tau_paths)

best_cp_paths <- 
  reduce(list(paths_0cp, paths_1cp, paths_2cp), left_join) %>% 
    rename(start_age = x) %>% 
    pivot_longer(cols = starts_with(c("path", "bic")), names_to = c(".value", "path_type"), names_sep = "_") %>% 
    group_by(code, start_age, sex) %>% 
    mutate(
      bic_rank = as.character(rank(bic))
    ) %>% 
    ungroup() %>% 
    unnest(paths) %>% 
    left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  filter(!is.na(country)) %>% 
  group_by(country, sex, start_age) %>% 
  filter(bic_rank == 1) 
```

The best changepoint model (if any), along with the points, are shown for life expectancy at birth below

```{r}
best_cp_paths %>%
  filter(start_age == 0) %>% 
  ggplot(aes(x, y)) + 
  geom_path(colour = "red") + 
  facet_grid(sex ~ country ) + 
  expand_limits(y  = 0) + 
  geom_hline(yintercept = 0) +
  theme(
      axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
      legend.position = 'none'
  ) +
  geom_point(
    aes(year, delta_ex),
    inherit.aes = FALSE,
    data = change_in_ex_selected_countries %>% 
      filter(x == 0) %>% 
      filter(year >= 1979) %>% 
      left_join(country_code_lookup) %>% 
      mutate(
        country = factor(
          country, 
          levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands")
          )
        ) %>% 
        filter(!is.na(country))   %>% 
        filter(between(year, 1980, 2020))  
  ) + 
  labs(
    x = "Year", 
    y = "Change in life expectancy since previous year",
    title = "Best changepoint model fits: Life expectancy at birth",
    subtitle = "Polyline indicates fit of best performing model",
    caption = "Source: HMD"
  )

We can see that the best-performing model is a two changepoing model with contiguous years for Scottish females, Spanish females, French females, Italian females, French males and Dutch males. For each of these models it is clear that the model in effect 'detected' an outlier of either exceptionally high annual change (France and Italy) or exceptionally low or negative annual change (Scottish males, Spanish females, Dutch males).

For England & Wales (both males and females) and for Scottish females, a single changepoint model was selected, with lower rates of annual life expectancy gain after around 2011 than before. For Scottish males this fell to negative values (declining rather than stalling mortality). A somewhat similar pattern is observed for Germany, but this appears to be less consistent in term and less severe than for England & Wales.

The equivalent graph for life expectancy from age 65 is shown below:

```{r}
best_cp_paths %>%
  filter(start_age == 65) %>% 
  ggplot(aes(x, y)) + 
  geom_path(colour = "red") + 
  facet_grid(sex ~ country ) + 
  expand_limits(y  = 0) + 
  geom_hline(yintercept = 0) +
  theme(
      axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
      legend.position = 'none'
  ) +
  geom_point(
    aes(year, delta_ex),
    inherit.aes = FALSE,
    data = change_in_ex_selected_countries %>% 
      filter(x == 65) %>% 
      filter(year >= 1979) %>% 
      left_join(country_code_lookup) %>% 
      mutate(
        country = factor(
          country, 
          levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands")
          )
        ) %>% 
        filter(!is.na(country))   %>% 
        filter(between(year, 1980, 2020))  
  ) + 
  labs(
    x = "Year", 
    y = "Change in life expectancy since previous year",
    title = "Best changepoint model fits: Life expectancy at age 65 years",
    subtitle = "Polyline indicates fit of best performing model",
    caption = "Source: HMD"
  )

```
For changes in conditional life expectancy from age 65 two changepoint models with contiguous years were identified for females in Germany, France and Italy; and for males in Scotland, France, Italy and the Netherlands.

# Summary and conclusion

This appendix has show the implementation and results of applying an alternative method to identifying distinct changes in trends in life expectancy for the selected countries. In many cases, a two changepoint model was identified. However, often these two changepoint models involved contiguous years, and appear to in effect be operating as outlier detection algorithms, picking out exceptional years from the series.

Although the fact these individual years are outliers is important to note, as it illustrates how varied life expectancy change series can be when single years are used, these single years cannot, by definition, define or constitute a change in trend, which was the purpose of this exercise. It is for this reason that the segmented approach was used instead of this approach in the main paper.

The changepoint approach could be adapted to be less 'sensitive' to outlier years by narrowing the search space for $tau_2$ conditional on $tau_1$ so as to be two or more years later. (e.g. if $tau_1$ were 1990 then candidate values for $tau_2$ would start from 1992 instead of 1991.) However, the choice of how much to narrow down the search space in this way is somewhat arbitrary, so was not adapted further for this paper.
